%=================================================================
% DRAFT MANUSCRIPT -- work in progress, section by section.
% Mirrors the structure/format of futureinternet-4330203-done (1).tex
% (Nouman, Ujjan & Hassanu, MDPI Future Internet), per the same
% supervisor's template, adapted to this project's own dataset,
% methods, and results (see artifacts/paper/tables and thesis_context
% memory for the underlying numbers).
%
% NOTE: Definitions/mdpi.cls is NOT yet present in this repo (the peer
% .tex file references it but the class files weren't included when
% the reference files were added). This document cannot be compiled
% until that class is sourced from MDPI's submission template. Content
% is being drafted in full MDPI-command form now so nothing needs to
% be rewritten once the class is available -- only the \documentclass
% line and packages need the real class file to render.
%=================================================================
\documentclass[futureinternet,article,accept,moreauthors]{Definitions/mdpi}

\firstpage{1}
\makeatletter
\setcounter{page}{\@firstpage}
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2026}
\copyrightyear{2026}
\externaleditor{Firstname Lastname}
\datereceived{ }
\daterevised{ }
\dateaccepted{ }
\datepublished{ }
\hreflink{https://doi.org/}

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{float}
\usepackage{tabularx}
\usepackage{changepage}
\usepackage{array}
\usepackage{ragged2e}
\usepackage[T1]{fontenc}
\usepackage{booktabs}

%=================================================================
% TITLE -- DRAFT, needs confirmation. Mirrors the peer paper's
% "[qualifier] Anomaly Detection for [traffic] Using [method A] with
% [method B]" pattern.
\Title{Content- and Session-Aware Anomaly Detection for Web Server Logs Using a Fused Autoencoder and Variational LSTM Framework: A Cross-Dataset Evaluation}

% AUTHOR BLOCK -- DRAFT. Matches the peer paper's author-role pattern
% (student first/corresponding author, supervisor + moderator as
% co-authors). NOTE: the moderator's name is spelled inconsistently
% across source documents ("Muhsin Husseinu" on the thesis title page,
% "Muhsin Hussenu" in the thesis body/acknowledgements, "Muhsin
% Hassanu" in the peer paper) -- needs to be confirmed with the
% moderator directly before submission, not guessed.
\Author{Abdul Rehman *, Raja Ujjan and Muhsin Hussenu *}

\AuthorNames{Abdul Rehman, Raja Ujjan, and Muhsin Hussenu}

\address[1]{Department of Computing and Cybersecurity, University of the West of Scotland, London Campus, London E14 2BA, UK}

\corres{Correspondence: [student email TBD] (A.R.)}

\abstract{%TODO: write after Results/Discussion are finalized, so the
reported numbers are locked. Draft note: should mirror the peer
paper's abstract shape -- motivation, gap, framework summary,
datasets, headline quantitative results, one-sentence takeaway.
Numbers to pull from once ready: artifacts/paper/tables/ablation_table.csv,
fusion_sweep.csv, threshold_transfer.csv, baseline_comparison.csv,
latency_benchmark.json, session_v1_vs_v2.csv (source of truth --
not the original thesis Table 5.6 figures, several of which this
re-run diverges from).}

\keyword{web server logs; anomaly detection; autoencoder; variational LSTM; session modelling; score fusion; cross-dataset generalisation}

\begin{document}

\section{Introduction}

Web applications underpin most critical digital services in use today, from banking and healthcare to e-commerce and cloud infrastructure, and their continuous, public-facing exposure makes them a persistent target for attackers~\cite{scarfone2007}. Every request a web server processes is recorded in an access log, together with metadata such as the request method, path, response status, and client identifiers. These logs are one of the richest available sources of security-relevant information: in principle, they capture both the legitimate use of an application and the traces left by attackers probing or exploiting it. In practice, however, the volume and variety of this data make manual inspection infeasible for any application of meaningful scale, and organisations have long relied on signature-based intrusion detection and rule-based filtering to triage it~\cite{scarfone2007}.

Signature- and rule-based approaches are effective against known attack patterns but are structurally unable to detect zero-day or previously unseen attacks, since they can only recognise what has already been catalogued. This has motivated a shift towards anomaly detection, which instead models what \emph{normal} traffic looks like and flags deviations from it, without requiring the deviation to match a predefined signature~\cite{chandola2009}. Anomaly detection is a natural fit for web security because it does not depend on a labelled corpus of known attacks -- which is difficult to obtain, quickly goes stale, and can never be assumed complete -- but it introduces its own difficulties. Web traffic is highly non-stationary: normal behaviour drifts as applications change, and it can vary enormously between different endpoints, user populations, and even times of day. Attack traffic is also typically a tiny fraction of the total volume, so any evaluation of an anomaly detector must be designed around severe class imbalance rather than raw accuracy~\cite{chandola2009}.

Deep learning has extended anomaly detection to log and web-traffic data by learning representations directly from raw or lightly processed records, rather than requiring hand-engineered features. Sequence models such as LSTMs have been used to learn the normal structure of a stream of log events and flag deviations from it~\cite{nedelkoski2020}, and DeepLog demonstrated that treating a system log as a natural-language-like sequence allows an LSTM to learn normal execution patterns and detect anomalies as departures from them~\cite{du2017}. Applied to HTTP traffic specifically, this behavioural, sequence-oriented view is complementary to a second, request-level view: individual HTTP requests can themselves be modelled directly, using an autoencoder trained to reconstruct the structural and payload characteristics of normal requests, so that unusually large reconstruction error flags anomalous content such as injection payloads or malformed parameters. Recent work combining a variational LSTM autoencoder with a deviation network for exactly this purpose has shown that this class of model is well suited to detecting web attacks directly from HTTP weblogs~\cite{jagat2024}.

These two views, however, are not interchangeable. A content-based model that scores individual requests in isolation cannot see attacks that only become apparent across a sequence of otherwise unremarkable requests -- reconnaissance-style scanning, credential stuffing, or low-and-slow probing, for example, where each individual request may be syntactically valid. A behaviour-based model that scores sequences of requests grouped by client can catch exactly this class of attack, but it depends on there being a genuine sequence to model in the first place: if a client only issues one or two requests, or if requests from a given client cannot be reliably grouped together, there is no behavioural signal to learn from. This dependency on session structure is not merely a theoretical caveat -- as this study's own experiments later show (Section~\ref{sec5}), it materially determines whether a session-based branch can contribute anything at all for a given traffic sample.

Because content-based and behaviour-based detectors fail in different, largely non-overlapping ways, combining their outputs is a natural way to cover more of the attack surface than either can alone. Hybrid, fusion-based frameworks that integrate multiple detection signals have been shown to improve robustness and stability relative to single-model detectors, particularly under the kind of severe class imbalance and limited labelled data that characterises real security settings~\cite{kamal2025}. Realising this benefit in practice, however, depends heavily on how the resulting system is evaluated. A large share of the anomaly-detection literature reports results on a single dataset, which risks overstating how well a method will generalise once deployed against traffic with different structure, format, or attack composition than whatever it was tuned against. This concern is well established in the wider intrusion-detection literature, where the design of benchmark datasets themselves -- their realism, diversity, and reproducibility -- has long been recognised as a determining factor in whether reported results are meaningful outside the dataset they were produced on~\cite{shiravi2012}.

Despite the individual maturity of content-based autoencoders, behaviour-based sequence models, and fusion-based detection frameworks, their combined application to web server logs specifically, evaluated across genuinely different log formats and under a strict cross-dataset protocol -- where a detection threshold is fixed on one data source and applied unchanged to others, rather than recalibrated on each dataset in turn -- remains underexplored. Many studies that do combine content and behavioural signals validate the result on a single log format and a single traffic sample, leaving open how such a system behaves when the underlying session structure, log schema, or class balance of the target traffic differs substantially from what the model was built on.

To address this gap, this work presents a hybrid anomaly detection framework for web server logs that combines a content-based autoencoder, trained to reconstruct the structural features of individual HTTP requests, with a session-based variational LSTM autoencoder, trained on genuine per-client request sequences rather than an arbitrary sliding window over the raw log stream. The outputs of the two branches are combined through score-level fusion, and the resulting framework is evaluated not only within a single dataset but across three structurally different sources: two self-generated Apache-style access-log corpora and a structured Nginx JSON log, together with the publicly available CSIC~2010 HTTP dataset, which additionally serves as the source domain for a cross-dataset threshold-transfer evaluation in which a single fixed decision threshold, calibrated once on genuinely benign traffic, is applied unchanged to unseen target datasets.

The remainder of this paper is organised as follows. Section~\ref{sec2} sets out the specific contributions of this work. Section~\ref{sec3} reviews related work on anomaly detection for web and log data, content- and behaviour-based deep learning models, and hybrid/fusion-based detection frameworks. Section~\ref{sec4} describes the datasets, preprocessing, feature representation, model architectures, and evaluation protocol. Section~\ref{sec5} presents the experimental results, including the branch ablation, fusion-weight sensitivity, classical-baseline comparison, and cross-dataset threshold-transfer experiments. Section~\ref{sec6} discusses these findings and their practical implications, and Section~\ref{sec7} concludes the paper and outlines directions for future work.

\section{Research Contributions} \label{sec2}

This paper makes the following contributions.

First, we build a hybrid system that finds anomalies in web server logs using two parts. One part checks each request on its own, using an autoencoder. The other part checks a sequence of requests from the same client, using a variational LSTM. We combine the two scores into one final score, so the system can catch both strange single requests and strange patterns of behaviour over time.

Second, we find and fix a real problem in how the behaviour part was trained. In the old version, the model was trained using a plain sliding window over the whole log. This window did not check which client sent each request, so requests from different clients could end up mixed in the same window. We fix this by grouping requests by client (IP address and user agent) first, so each window only holds requests from one real client. This fix makes a big difference: on the CSIC~2010 dataset, where one client sends many requests in a row, the behaviour part's AUC score rises from about 0.63 to about 0.97 after the fix.

Third, this work reports an honest and useful finding about the data itself. Some of our datasets have only one request per client. In real per-session detection, if a client only sends one request, there is no sequence to check, so the behaviour part gives no score at all for that data. We treat this as a fact about the data, not a failure of the method, and we explain it clearly instead of hiding it. This shows that a true session-based model only helps when the traffic actually has multi-request sessions in it.

Fourth, we test the system on more than one kind of log. We use two Apache-style logs we built ourselves, one Nginx JSON log, and the public CSIC~2010 HTTP dataset. Testing across different log formats and different datasets shows whether the method still works when the data changes shape.

Fifth, we run a cross-dataset test that checks if a threshold set on one dataset still works on another. We set the detection threshold using only the CSIC data, and only its normal, non-attack traffic. Then we use that same threshold, with no change, on three other datasets. The result is surprising: F1 score goes up, not down, when we use the frozen CSIC threshold instead of a threshold tuned on each dataset on its own. For example, on the Access Eval Mix 2000 dataset, the fused F1 score rises from about 0.12 (tuned on that dataset) to about 0.67 (using the frozen CSIC threshold). This happens because these test datasets contain a lot of attack traffic, so a threshold tuned only on them is not a fair "normal traffic" threshold. This is a useful, real finding about how to set thresholds in practice.

Sixth, we compare our system with simple baseline methods: Random Forest, which uses attack labels during training, and Isolation Forest and One-Class SVM, which do not use labels, like our own method. We also measure how fast the system runs. On average, scoring one request takes about 0.14~ms for the content part and about 0.08~ms for the behaviour part when done one request at a time, and scoring in batches is much faster, at about 630{,}000 lines per second. This shows the system could run in near real time.

Together, these results show that a two-part fusion system can work well across different kinds of web log data, and that fixing how sessions are built and how thresholds are set can matter as much as the choice of model itself.

\section{Related Work} \label{sec3}

\subsection{Anomaly Detection in Web Server Logs}

Web server logs record every request a server handles. Each line usually has the request method, the URL, the response code, the time, and the client's IP address. Security teams can use these logs to find attacks, but there are far too many lines to check by hand. For a long time, most systems used rules or signatures to spot bad requests~\cite{scarfone2007}. Rules like this work well for known attacks, but they cannot catch new or changed attacks, because they only look for patterns someone has already written down. This is why anomaly detection is useful: instead of matching known attack patterns, it learns what normal traffic looks like and flags anything that looks different~\cite{chandola2009}. This can catch attacks that have never been seen before, but it is harder to build well, because normal web traffic changes over time and attacks are a very small part of all traffic.

\subsection{Classical Machine Learning Methods}

Before deep learning became common, many studies used classic machine learning methods for this problem. Isolation Forest is one of the most used. It works by randomly splitting the data again and again; points that get cut off quickly, after only a few splits, are treated as anomalies~\cite{liu2008}. This method does not need labelled attack data, runs fast, and is simple to set up, which is why it is still used today. For example, one recent study applied Isolation Forest directly to web traffic logs from an e-commerce site and showed it can tell normal and anomalous requests apart without needing attack labels~\cite{chua2024}. Other classic methods, such as clustering and one-class classifiers, have also been used, but they often need features to be picked and built by hand, and they can struggle when the traffic pattern changes.

\subsection{Content-Based Deep Learning}

Deep learning removes the need for hand-built features by learning directly from the raw or lightly processed request data. One common approach is the autoencoder. An autoencoder learns to compress a normal request into a small representation and then rebuild it. If the model is trained only on normal requests, it becomes good at rebuilding normal requests but bad at rebuilding attack requests, so a request with a high rebuild error can be flagged as an anomaly. One study used this idea end to end, training a deep autoencoder directly on HTTP requests, and showed it could detect attacks such as SQL injection and cross-site scripting without needing much labelled data~\cite{pan2019}. Models like this are good at finding strange or broken single requests, but they look at each request on its own, so they cannot see attacks that only show up across many requests over time.

\subsection{Behaviour-Based Deep Learning}

To catch attacks that unfold over several requests, other work models the order and pattern of requests instead of looking at single requests. LSTM networks are built for this, since they can learn from a sequence of past events. DeepLog treats a stream of log lines like a sentence, learning the normal order of events and flagging any sequence that breaks the learned pattern~\cite{du2017}. Later work used a self-attention model, instead of or alongside an LSTM, to learn which parts of a log sequence matter most for telling normal and abnormal apart~\cite{nedelkoski2020}. Closer to our own work, one study built a variational LSTM autoencoder combined with a deviation network, trained directly on HTTP weblogs, and showed it can detect web attacks by modelling normal session behaviour~\cite{jagat2024}. Behaviour-based models like these can catch attacks that content-based models miss, such as scanning or repeated probing, but they need enough real, connected requests from the same client to learn from. If a client only sends one or two requests, there is no real sequence to learn from -- a point this paper returns to directly in Section~\ref{sec5}.

\subsection{Hybrid and Fusion-Based Detection}

Since content-based and behaviour-based models catch different kinds of attacks, several studies combine both into one system. The idea is to score each request or session with more than one model, then merge the scores into a single final decision. This kind of hybrid, fusion-based design has been shown to work better than either model alone, especially when labelled attack data is scarce and normal traffic looks very different from attack traffic~\cite{kamal2025}. Building a good fusion system is not simple, though: the two branches usually need to be trained and set up carefully so that one does not drown out the other, and the merged score still needs its own threshold.

\subsection{Cross-Dataset Evaluation}

Most of the studies above are tested on a single dataset. This is a real problem, because a model that works well on the data it was tuned on may not work as well on new data with a different shape, format, or mix of attacks. Building good benchmark datasets for intrusion detection, and testing across more than one of them, has long been seen as important for getting results that mean something outside the exact dataset used~\cite{shiravi2012}. A recent study looked at this problem directly for network intrusion detection: models that scored close to perfect when trained and tested on the same dataset dropped close to random guessing when tested on a different dataset, showing how easy it is to overstate a model's real performance with single-dataset testing~\cite{cantone2024}. This is the exact gap this paper addresses for web server logs: testing across more than one log format, and running a true cross-dataset threshold test, where the decision threshold is set once and never changed for new data.

\subsection{Summary}

Table~\ref{tab:lit_review} lists the studies discussed above, the kind of method each one uses, the kind of data it was tested on, and the main gap that this paper tries to close.

\begin{table}[H]
\caption{Summary of related studies on web and log anomaly detection.}
\label{tab:lit_review}
\scriptsize
\begin{adjustwidth}{-\extralength}{0cm}
\begin{tabularx}{\fulllength}{l l l l m{4.5cm}<{\raggedright}}
\toprule
\textbf{Study} & \textbf{Method} & \textbf{Data Type} & \textbf{Evaluation} & \textbf{Main Limitation} \\
\midrule

Scarfone and Mell~(2007)~\cite{scarfone2007} & Rule-based IDS/IPS guidance & Network traffic & Guidance document & Rules only catch known attack patterns \\

Chandola et al.~(2009)~\cite{chandola2009} & Survey & General & Review study & No single detector proposed \\

Liu et al.~(2008)~\cite{liu2008} & Isolation Forest & General tabular data & Benchmark datasets & Not evaluated on web request logs \\

Chua et al.~(2024)~\cite{chua2024} & Isolation Forest & Web traffic logs & Single dataset & Single detection view, no behaviour model \\

Pan et al.~(2019)~\cite{pan2019} & Stacked denoising autoencoder & HTTP requests & Single dataset & Scores single requests only, no session view \\

Du et al.~(2017)~\cite{du2017} & LSTM (DeepLog) & System logs & Single dataset & Not web-request specific \\

Nedelkoski et al.~(2020)~\cite{nedelkoski2020} & Self-attention classifier & Unstructured logs & Single dataset & No cross-dataset test reported \\

Jagat et al.~(2024)~\cite{jagat2024} & Variational LSTM autoencoder + deviation network & HTTP weblogs & Single dataset & No content-based branch, single dataset \\

Kamal and Mashaly~(2025)~\cite{kamal2025} & Hybrid deep learning IDS & Network traffic & Single dataset & Single dataset, not web-log specific \\

Shiravi et al.~(2012)~\cite{shiravi2012} & Benchmark dataset design & Network traffic & Dataset study & Dataset paper, no detection model \\

Cantone et al.~(2024)~\cite{cantone2024} & Multiple classical ML models & Network traffic & Cross-dataset & Not applied to web-log or hybrid deep models \\

\bottomrule
\end{tabularx}
\end{adjustwidth}
\end{table}
\unskip

\section{Materials and Methods} \label{sec4}
%TODO: next section to draft.

\section{Results} \label{sec5}
%TODO: next section to draft. Source data: artifacts/paper/tables/{ablation_table,fusion_sweep,threshold_transfer,baseline_comparison,session_v1_vs_v2}.csv, latency_benchmark.json.

\section{Discussion} \label{sec6}
%TODO: next section to draft.

\section{Conclusions} \label{sec7}
%TODO: next section to draft.

%=================================================================
\begin{adjustwidth}{-\extralength}{0cm}
\reftitle{References}
\begin{thebibliography}{999}

\bibitem[Scarfone and Mell(2007)]{scarfone2007}
Scarfone, K.; Mell, P. \emph{Guide to Intrusion Detection and Prevention Systems (IDPS)}; NIST Special Publication 800-94; National Institute of Standards and Technology: Gaithersburg, MD, USA, 2007. \url{https://doi.org/10.6028/NIST.SP.800-94}.

\bibitem[Chandola et al.(2009)]{chandola2009}
Chandola, V.; Banerjee, A.; Kumar, V. Anomaly detection: A survey. \emph{ACM Comput. Surv.} \textbf{2009}, \emph{41}, 15. \url{https://doi.org/10.1145/1541880.1541882}.

\bibitem[Nedelkoski et al.(2020)]{nedelkoski2020}
Nedelkoski, S.; Bogatinovski, J.; Acker, A.; Cardoso, J.; Kao, O. Self-Attentive Classification-Based Anomaly Detection in Unstructured Logs. In Proceedings of the 2020 IEEE International Conference on Data Mining (ICDM), Sorrento, Italy, 17--20 November 2020; pp. 1196--1201. \url{https://arxiv.org/abs/2008.09340}.

\bibitem[Du et al.(2017)]{du2017}
Du, M.; Li, F.; Zheng, G.; Srikumar, V. DeepLog: Anomaly Detection and Diagnosis from System Logs through Deep Learning. In Proceedings of the 2017 ACM SIGSAC Conference on Computer and Communications Security (CCS), Dallas, TX, USA, 30 October--3 November 2017; pp. 1285--1298.

\bibitem[Jagat et al.(2024)]{jagat2024}
Jagat, R.R.; Sisodia, D.S.; Singh, P. Detecting Web Attacks From HTTP Weblogs Using Variational LSTM Autoencoder Deviation Network. \emph{IEEE Trans. Serv. Comput.} \textbf{2024}, \emph{17}, 2210--2222. \url{https://doi.org/10.1109/TSC.2024.3453748}.

\bibitem[Kamal and Mashaly(2025)]{kamal2025}
Kamal, H.; Mashaly, M. Enhanced Hybrid Deep Learning Models-Based Anomaly Detection Method for Two-Stage Binary and Multi-Class Classification of Attacks in Intrusion Detection Systems. \emph{Algorithms} \textbf{2025}, \emph{18}, 69. \url{https://doi.org/10.3390/a18020069}.

\bibitem[Shiravi et al.(2012)]{shiravi2012}
Shiravi, A.; Shiravi, H.; Tavallaee, M.; Ghorbani, A.A. Toward developing a systematic approach to generate benchmark datasets for intrusion detection. \emph{Comput. Secur.} \textbf{2012}, \emph{31}, 357--374. \url{https://doi.org/10.1016/j.cose.2011.12.012}.

\bibitem[Liu et al.(2008)]{liu2008}
Liu, F.T.; Ting, K.M.; Zhou, Z.-H. Isolation Forest. In Proceedings of the 2008 Eighth IEEE International Conference on Data Mining (ICDM), Pisa, Italy, 15--19 December 2008; pp. 413--422. \url{https://doi.org/10.1109/ICDM.2008.17}.

\bibitem[Chua et al.(2024)]{chua2024}
Chua, W.; Pajas, A.L.D.; Castro, C.S.; Panganiban, S.P.; Pasuquin, A.J.; Purganan, M.J.; Malupeng, R.; Pingad, D.J.; Orolfo, J.P.; Lua, H.H.; et al. Web Traffic Anomaly Detection Using Isolation Forest. \emph{Informatics} \textbf{2024}, \emph{11}, 83. \url{https://doi.org/10.3390/informatics11040083}.

\bibitem[Pan et al.(2019)]{pan2019}
Pan, Y.; Sun, F.; Teng, Z.; White, J.; Schmidt, D.C.; Staples, J.; Krause, L. Detecting web attacks with end-to-end deep learning. \emph{J. Internet Serv. Appl.} \textbf{2019}, \emph{10}, 16. \url{https://doi.org/10.1186/s13174-019-0115-x}.

\bibitem[Cantone et al.(2024)]{cantone2024}
Cantone, M.; Marrocco, C.; Bria, A. Machine learning in network intrusion detection: A cross-dataset generalization study. \emph{IEEE Access} \textbf{2024}, \emph{12}, 144489--144508. \url{https://doi.org/10.1109/ACCESS.2024.3472907}.

\end{thebibliography}
\end{adjustwidth}

\end{document}
